\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-24 NAR_DRAFT.tex
Minor introduction edits

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity \citep{Koboldt2012,Cibulskis2013}.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, BAYSIC, SomaticSeq, and SNooPer) \citep{Ding2012,Cantarel2014,Fang2015,Spinella2016}.
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is constant across the genome.

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
The processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, Figure~\ref{NAR-realsigfig}, shows mutation profiles from several tumors.
These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
Our algorithm estimates the tumor's mutation profile using high-confidence variants and then uses that profile as a prior when calling other variants.
Our implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
Using both simulated and real data, we show that the addition of a mutation profile prior to MuTect produces a superior variant caller.
Our algorithm is simple, computationally inexpensive, and can be integrated into numerous other variant callers.
Broad adoption of our approach will enable more confident study of low allelic frequency mutations in tumors in both research and clinical settings.

\begin{figure}
  \begin{center}
  \includegraphics{figures/real_signatures_only.pdf}
  \end{center}
  \caption{Tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two breast tumors \cite{Alexandrov2019}.
  \rngcomment{Label cut off in panel A. Since this figure is now all real data, we can remove "real" from each internal label.}
   }
  \label{NAR-realsigfig}
 \end{figure}

>>>>

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1~\cite{Cibulskis2013} and Varscan2~\cite{Koboldt2012}, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, BAYSIC, SomaticSeq, and SNooPer) \citep{Ding2012,Cantarel2014,Fang2015,Spinella2016}.
\rngcomment{Connect citations to individual tools. (See MuTect1 and Varscan2 examples earlier in paragraph.}
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is uniform across the genome.

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
If fact, even the processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, the mutation profiles differs substantially among the three breast tumors illustrated in Figure~\ref{NAR-realsigfig}B-D.
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
Our algorithm first estimates the tumor's mutation profile using high-confidence variants and then uses that profile as a prior when calling other variants.
Our R implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
Using both simulated and real data, we show that the addition of a mutation profile prior to MuTect produces a superior variant caller.
Our algorithm is simple and computationally inexpensive, and it can be integrated into numerous other variant callers.
Broad adoption of our approach will enable more confident study of low allelic frequency mutations in tumors in both research and clinical settings.

\begin{figure}
  \begin{center}
  \includegraphics{figures/real_signatures_only.pdf}
  \end{center}
  \caption{Tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two additional breast tumors \cite{Alexandrov2019}.
  \rngcomment{Label cut off in panel A. Since this figure is now all real data, we can remove "real" from each internal label.}
   }
  \label{NAR-realsigfig}
 \end{figure}
